---
title: "Example: Spatial modeling of trees in a rainforest"
subtitle: "NOAA PSAW Seminar Series"
author: ""
institute: ""
date: "March 2, 2022"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

<!-- Build with: xaringan::inf_mr() -->

```{r preamble, include=FALSE, cache=FALSE}
source(here::here("noaa-psaw-2022/preamble.R"))
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(dplyr)
library(sdmTMB)
library(ggplot2)
```

# The bei dataset

.medium[
```{r, echo=TRUE}
dat <- data.frame(
  x = spatstat.data::bei$x, 
  y = spatstat.data::bei$y
)
head(dat)
```
]

---

# Spatial patterning of trees

* Barro Colorado Island = classic in ecology
  * Hubbell's (2001) unified neutral theory of biodiversity
  
```{r plot-trees, fig.asp=0.5}
ggplot(dat, aes(x, y)) +
  geom_point(col = "darkblue", alpha = 0.1) +
  coord_cartesian(expand = FALSE)
```

---

# Lattice representation

* Used in INLA book and other R packages ('spatstat')  

```{r plot-trees-grid, fig.asp=0.5}
# scale controls resolution
scale <- 50
dat$x <- scale * floor(dat$x / scale)
dat$y <- scale * floor(dat$y / scale)

dat <- dplyr::group_by(dat, x, y) %>%
  dplyr::summarise(n = n())

ggplot(dat, aes(x, y, fill = n)) +
  geom_tile() +
  theme_bw() +
  coord_fixed(expand = FALSE) +
  scale_fill_viridis_c()
```

---

# Should a spatial model be used?

* Diagnostics: Moran's I

.small[
```{r echo=TRUE}
library(ape)
inv_dists <- as.matrix(dist(dat[,c("x","y")]))
diag(inv_dists) <- 0
Moran.I(dat$n, inv_dists)
```
]  

???
Could also fit model using `gls()` to estimate spatial range, etc.
 
---

# What type of distribution?

* Counts
* Skewed
* Zero-truncated (locations of measured trees)

```{r tree-hist, fig.asp=0.5}
ggplot(dat, aes(n)) +
  geom_histogram(breaks = seq(1, 200), col = "darkblue") +
  theme_bw() +
  xlab("# Trees") +
  coord_cartesian(expand = FALSE, xlim = c(0, 130))
```

---

# What type of distribution?

* Likely best fit as a point process model for presence-only analysis (e.g., [Renner et al. 2015](https://doi.org/10.1111/2041-210X.12352))
  * could be done with a Poisson family and weights in sdmTMB
* For simplicity, we could ignore truncation and use `nbinom2()`
* sdmTMB has several additional families for censored data:
  * `truncated_nbinom1()`
  * `truncated_nbinom2()`
  * `censored_poisson()`
  
---

# Creating a mesh

.small[
```{r make-mesh, echo=TRUE, out.width='340px', fig.asp=0.8}
mesh <- make_mesh(
  dat, 
  xy_cols = c("x", "y"), 
  cutoff = 80 # min. distance between knots in X-Y units
)
mesh$mesh$n # extract number of vertices/knots

# quick visualization of mesh
plot(mesh)
```
]

---

# Fancier mesh visualizations

.small[
```{r fig.asp=0.5, echo=TRUE}
ggplot() +
  inlabru::gg(mesh$mesh) + #<<
  geom_point(data = dat, aes(x = x, y = y, col = n)) +
  coord_equal()
```
]

---

# A first model

```{r fit-model, echo=TRUE}
fit <- sdmTMB(
  n ~ 1,
  data = dat,
  mesh = mesh,
  family = truncated_nbinom2(link = "log"),
)
```

---

# Inspecting the model output

.small[
```{r inspect1, echo=TRUE, highlight.output=1}
summary(fit)
```
]

---

# Inspecting the model output

.small[
```{r inspect2, echo=TRUE, highlight.output=2:5}
summary(fit)
```
]

---

# Inspecting the model output

.small[
```{r inspect3, echo=TRUE, highlight.output=6:7}
summary(fit)
```
]

---

# Inspecting the model output

.small[
```{r inspect4, echo=TRUE, highlight.output=9:11}
summary(fit)
```
]

---

# Extracting parameters in a data frame

```{r extract, echo=TRUE}
# Main effects:
tidy(fit)

# Variance-related terms:
tidy(fit, effects = "ran_pars")
```

---

# Extracting parameters in a data frame

.small[
```{r, echo=TRUE}
tidy(fit, "ran_pars", conf.int = TRUE) %>%
  as_tibble()
```
]

---

# Making predictions on original data

* sdmTMB has a `predict()` function just like other packages
* For help, see `?predict.sdmTMB`

.small[
```{r, echo=TRUE}
p <- predict(fit)
select(p, x, y, est:omega_s) %>%
  as_tibble()
```
]

---

# What are the columns in the prediction output?

* `est`: Overall estimate **in link space** (logit here)
* `est_non_rf`: Estimate of non-random-field components
* `est_rf`: Estimate of random-field components
* `omega_s`: Spatial random field

---

# Making predictions on new data

* Let's project to the whole region (on a grid)
* Create a new data frame for predictions

.small[
```{r echo = TRUE}
# makes all combinations of x and y:
newdf <- expand.grid( 
  x = unique(dat$x),
  y = unique(dat$y)
)
head(newdf)
```
]

---

# Making predictions on new data

* `predict.sdmTMB()` can also take `newdata`:

```{r, echo=TRUE}
p <- predict(fit, newdata = newdf)
```

See `?predict.sdmTMB`


---

# Plotting fixed effects (the intercept here)

.small[
```{r, echo=TRUE}
ggplot(p, aes(x, y, fill = est_non_rf)) + #<<
  geom_raster() +
  coord_fixed(expand = FALSE)
```
]

---

# Plotting spatial random effects

.small[
```{r, echo=TRUE}
ggplot(p, aes(x, y, fill = omega_s)) + #<<
  geom_raster() +
  scale_fill_gradient2() +
  coord_fixed(expand = FALSE)
```
]

---

# Predictions in link (log) space

.xsmall[
```{r, echo=TRUE}
ggplot(p, aes(x, y, fill = est)) + #<<
  geom_raster() + scale_fill_viridis_c() +
  coord_fixed(expand = FALSE) +
  geom_point(aes(x, y, size = n), 
    data = dat, pch = 21, colour = "grey60", inherit.aes = FALSE)
```
]

---

# Predictions in natural space


.xsmall[
```{r, echo=TRUE}
# Hint: inverse link function also available in `fit$family$linkinv()`
ggplot(p, aes(x, y, fill = exp(est))) + #<<
  geom_raster() + scale_fill_viridis_c() +
  coord_fixed(expand = FALSE) +
  geom_point(aes(x, y, size = n), 
    data = dat, pch = 21, colour = "grey60", inherit.aes = FALSE)
```
]

