---
title: "Example: Spatial modeling of Pacific cod"
subtitle: "NOAA PSAW Seminar Series"
author: ""
institute: ""
date: "2022-03-02"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

<!-- Previous slides: https://www.dropbox.com/s/gu43n4gvqmo0uzy/sdmTMB-intro-2021.pdf?dl=1 -->
<!-- Build with: xaringan::inf_mr() -->

```{r preamble, include=FALSE, cache=FALSE}
source(here::here("noaa-psaw-2022/preamble.R"))
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(dplyr)
library(sdmTMB)
library(ggplot2)
```

---

class: center, middle, inverse

# A basic spatial example

---

# The Pacific Cod dataset

DFO trawl survey data from Queen Charlotte Sound off BC

.small[
```{r, echo=TRUE}
library(dplyr)
library(ggplot2)
library(sdmTMB)
pcod %>%
  select(year, X, Y, depth, density, present) %>% 
  head()
```
]

---

# Building a mesh

.small[
```{r, echo=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
plot(mesh)
```
]

---

# Adding spatial random fields

```{r pcod-spatial-rf, echo=TRUE}
fit <- sdmTMB(
  present ~ s(depth),
  data = pcod,
  mesh = mesh,
  family = binomial(link = "logit"),
  spatial = "on"   #<<
)
```

---

# Inspecting the model output

.small[
```{r, echo=TRUE, highlight.output=c(1)}
fit
```
]

---

# Inspecting the model output

.small[
```{r, echo=TRUE, highlight.output=2:5}
fit
```
]

---

# Inspecting the model output

.small[
```{r, echo=TRUE, highlight.output=6:9}
fit
```
]

---

# Inspecting the model output

.small[
```{r, echo=TRUE, highlight.output=11:13}
fit
```
]

---

# Extracting parameters in a data frame

```{r, echo=TRUE}
# Main effects:
tidy(fit)

# Variance-related terms:
tidy(fit, effects = "ran_pars")
```

---

# Extracting parameters in a data frame

.small[
```{r, echo=TRUE}
tidy(fit, "ran_pars", conf.int = TRUE) %>% 
  as_tibble()
```
]

---

# Making predictions

.small[
```{r, echo=TRUE}
p <- predict(fit) #<<
select(p, X, Y, est:omega_s) %>% 
  as_tibble()
```
]

---

# Making predictions

* `est`: Overall estimate **in link space**
* `est_non_rf`: Estimate of non-random-field components
* `est_rf`: Estimate of random-field components
* `omega_s`: Spatial random field

---

# Making predictions on new data

.small[
* `qcs_grid`: a 2x2 km grid extending over the full survey domain
]
.small[
```{r, echo=TRUE, out.width="500px", fig.width=5}
nd <- filter(qcs_grid, year == 2017) # pick any year
ggplot(nd, aes(X, Y, fill = depth)) + 
  geom_raster() +
  coord_fixed()
```
]

---

# Making predictions on new data


```{r, echo=TRUE}
head(select(nd, X, Y, depth))

p <- predict(fit, newdata = nd) #<<
```

---

# Plotting predictions on new data

.small[
```{r, echo=TRUE, fig.width=6, out.width="600px"}
ggplot(p, aes(X, Y, fill = plogis(est))) +
  geom_raster() +
  scale_fill_viridis_c() +
  coord_fixed()
```
]

---

# Plotting main effect contributions

.small[
```{r, echo=TRUE, fig.width=6, out.width="600px"}
ggplot(p, aes(X, Y, fill = plogis(est_non_rf))) +
  geom_raster() +
  scale_fill_viridis_c() +
  coord_fixed()
```
]

---

# Plotting spatial random effects

.small[
```{r, echo=TRUE, fig.width=6, out.width="600px"}
ggplot(p, aes(X, Y, fill = omega_s)) +
  geom_raster() +
  scale_fill_gradient2() +
  coord_fixed()
```
]

---

# Plotting depth effect

.small[
```{r pcod-sp-depth-pred, echo=TRUE}
nd <- data.frame(depth = seq(30, 300, length.out = 100))
p <- predict(
  fit, 
  newdata = nd, 
  se_fit = TRUE, #<<
  re_form = ~ 0 #<<
)
p$lwr <- plogis(p$est - 1.96 * p$est_se)
p$upr <- plogis(p$est + 1.96 * p$est_se)
```

```{r pcod-sp-depth-pred-plot, fig.width=5, echo=FALSE, out.width='500px', fig.asp=0.5}
ggplot(p, aes(nd$depth, plogis(est), ymin = lwr, ymax = upr)) +
  geom_ribbon(alpha = 0.2) +
  geom_line() + xlab("Depth (m)") + ylab("") +
  coord_cartesian(ylim = c(0, 1), expand = FALSE)
```
]

---

# Extending to a spatiotemporal model

Reminder: the Pacific Cod dataset:

.small[
```{r, echo=TRUE}
pcod %>% 
  select(year, X, Y, depth, density, present) %>% 
  head()
```
]

---

# Our previous spatial model fit

```{r pcod-spatial-rf-reminder, echo=TRUE, eval=FALSE}
fit <- sdmTMB(
  present ~ s(depth),
  data = pcod,
  mesh = mesh,
  family = binomial(link = "logit"),
  spatial = "on"
)
```

---

# Switching to density

```{r pcod-spatial-rf-density, echo=TRUE, eval=FALSE}
fit <- sdmTMB(
  density ~ s(depth), #<<
  data = pcod,
  mesh = mesh,
  family = tweedie(link = "log"), #<<
  spatial = "on"
)
```

---

# Adding spatiotemporal fields

```{r pcod-st-fit, echo=TRUE, eval=TRUE, results='hide'}
fit <- sdmTMB(
  density ~ s(depth) + 0 + as.factor(year), #<<
  data = pcod,
  mesh = mesh,
  family = tweedie(link = "log"),
  spatial = "on",
  time = "year", #<<
  spatiotemporal = "iid", #<<
  silent = FALSE # show progress!
)
```

---

# Inspecting the model fit

.small[
```{r, eval=FALSE, echo=TRUE}
fit
```

```{r pcod-st-print2, echo=TRUE, eval=TRUE}
#> Spatiotemporal model fit by ML ['sdmTMB']
#> Formula: density ~ s(depth) + 0 + as.factor(year)
#> Time column: "year"
#> Mesh: mesh
#> Data: pcod
#> Family: tweedie(link = 'log')
#>                     coef.est coef.se
#> as.factor(year)2003     1.80    0.28 #<<
#> as.factor(year)2004     2.40    0.27 #<<
#> ...
#> 
#> Dispersion parameter: 10.73
#> Tweedie p: 1.49
#> Matern range: 12.23 #<<
#> Spatial SD: 1.85
#> Spatiotemporal SD: 1.84 #<<
#> ML criterion at convergence: 6242.801
```
]

---

# Predicting on the survey grid

```{r pcod-st-pred, echo=TRUE}
p <- predict(fit, newdata = qcs_grid)
```

* `est`: Overall estimate **in link space**
* `est_non_rf`: Estimate of non-random-field components
* `est_rf`: Estimate of random-field components
* `omega_s`: Spatial random field
* `epsilon_st`: **Spatiotemporal random field**

---

# Plotting overall predictions

.xsmall[
```{r pcod-st-plot-est, echo=TRUE, fig.width=7}
ggplot(p, aes(X, Y, fill = exp(est))) +
  geom_raster() +
  scale_fill_viridis_c(trans = "sqrt") +
  facet_wrap(~year) +
  coord_fixed()
```
]

---

# Plotting main effect contributions

.xsmall[
```{r pcod-st-plot-non-rf, echo=TRUE, fig.width=7}
ggplot(p, aes(X, Y, fill = exp(est_non_rf))) +
  geom_raster() +
  scale_fill_viridis_c(trans = "sqrt") +
  facet_wrap(~year) +
  coord_fixed()
```
]

---

# Plotting spatial random effects

.xsmall[
```{r pcod-st-plot-omega, echo=TRUE, fig.width=7}
ggplot(p, aes(X, Y, fill = omega_s)) +
  geom_raster() +
  scale_fill_gradient2() +
  facet_wrap(~year) +
  coord_fixed()
```
]

---

# Plotting spatiotemporal random effects

.xsmall[
```{r pcod-st-plot-eps, echo=TRUE, fig.width=7}
ggplot(p, aes(X, Y, fill = epsilon_st)) +
  geom_raster() +
  scale_fill_gradient2() +
  facet_wrap(~year) +
  coord_fixed()
```
]

---

# Calculating spatial uncertainty

.small[
```{r pcod-st-sims, echo=TRUE}
psims <- predict(fit, newdata = qcs_grid, nsim = 500) #<<
dim(psims)
psims[1:3, 1:4]

p$sd <- apply(psims, 1, sd) #<<
```
]

---

# Calculating spatial uncertainty

.small[
```{r pcod-st-sims-plot, echo=TRUE, fig.width=5, out.width="500px"}
filter(p, year == 2017) %>% 
  ggplot(aes(X, Y, fill = sd)) +
  geom_raster() +
  scale_fill_viridis_c(trans = "log10") +
  coord_fixed()
```
]

---

# Calculating an area-weighted population index

.small[
```{r pcod-st-index1, echo=TRUE, fig.width=6, fig.asp=0.5, out.width="600px"}
p2 <- predict(fit, return_tmb_object = TRUE, area = 4) #<<
index <- get_index(p2, bias_correct = FALSE) #<<
ggplot(index, aes(year, est)) + geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
  xlab('Year') + ylab('Biomass estimate (kg)')
```
]

---

# Calculating an area-weighted population index

.small[
```{r pcod-st-index-sims, echo=TRUE, fig.width=6, fig.asp=0.5, out.width="600px"}
psims <- predict(fit, nsim = 1000) #<<
index_sims <- get_index_sims(psims, area = rep(4, nrow(psims))) #<<
ggplot(index_sims, aes(year, est)) + geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
  xlab('Year') + ylab('Biomass estimate (kg)')
```
]

---

# Calculating the center of gravity

.small[
```{r pcod-st-cog, echo=TRUE, fig.width=6, out.width="600px"}
cog <- get_cog(p2, bias_correct = FALSE, format = "wide") #<<
ggplot(cog, aes(est_x, est_y, colour = year)) + 
  geom_linerange(aes(xmin = lwr_x, xmax = upr_x)) +
  geom_linerange(aes(ymin = lwr_y, ymax = upr_y)) +
  coord_equal()
```
]


