---
title: "Troubleshooting, other packages, tips and tricks"
subtitle: "NOAA PSAW Seminar Series"
author: "Names Here"
institute: ""
date: "2022-03-02"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

<!-- Previous slides: https://www.dropbox.com/s/gu43n4gvqmo0uzy/sdmTMB-intro-2021.pdf?dl=1 -->
<!-- Build with: xaringan::inf_mr() -->

```{r preamble, include=FALSE, cache=FALSE}
source(here::here("noaa-psaw-2022/preamble.R"))
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(sdmTMB)
```



---

# Getting help

---

---


# Words of wisdom

---

---

# Related modelling software

* Fig from paper?
* Highlight table from paper?

---

# SPDE approach / INLA / etc; comparison with other software

<!-- --- -->

<!-- # Fitting basic spatial models in sdmTMB (Philina) -->

<!-- * Making meshes … link to INLA tutorial (+interactive meshbuilder) -->
<!-- * How do you know your model hasn’t converged? Who is Hessian and why hasn’t he converged? -->
<!-- * What warnings do you need to worry about?  -->
<!-- * Overall magnitude of errors / observation error etc.  -->

---

class: center, middle, inverse

# Troubleshooting

---

# Who is Hessian and why are they not "positive definite"??

* A Hessian matrix is a matrix of second derivatives of a function (here the negative log likelihoood surface)
* The inverse of the negative Hessian is the covariance matrix
* Warning means the curvature of the negative log-likelihood surface is inconsistent with the model having found the best fit.
* Overparameterized? Variance estimated near zero?
* See [vignette("troubleshooting", "glmmTMB")](https://cran.r-project.org/web/packages/glmmTMB/vignettes/troubleshooting.html)

---

# Splines

---

# Spatiotemporal models, different structures

---

# Prediction (prediction in time – e.g. future SDMs, prediction in space – new areas / unsampled areas)

---

---

# Fitting a basic GLM

.small[
```{r pcod-glm, echo=TRUE}
fit_glm <- glm(
  present ~ 1,
  data = pcod,
  family = binomial(link = "logit")
)

mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
fit <- sdmTMB(
  present ~ 1,
  data = pcod,
  mesh = mesh,
  family = binomial(link = "logit"),
  spatial = "off"
)
```
]
