---
title: "Spatially varying coefficients with sdmTMB"
subtitle: "NOAA PSAW Seminar Series"
author: ""
institute: ""
date: "March 9, 2022"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

```{r preamble, include=FALSE, cache=FALSE}
source(here::here("noaa-psaw-2022/preamble.R"))
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(dplyr)
library(sdmTMB)
library(ggplot2)
```

# What is a spatially varying coefficient?

---

# When might we want a spatially varying coefficient?

---

# Ecography paper figure here...

---

# Spatially varying coefficients in sdmTMB

Spatially varying effect of time:

```{r spat-varying-fit, echo=TRUE}
pcod$year_scaled <- as.numeric(scale(pcod$year))
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
fit <- sdmTMB(
  density ~ s(depth, k = 5) + year_scaled,
  spatial_varying = ~ 0 + year_scaled, #<<
  data = pcod, mesh = mesh, 
  time = "year",
  family = tweedie(link = "log"),
  spatiotemporal = "off"
)
```

<!-- See `zeta_s` in the output, which represents the coefficient varying in space. You'll want to ensure you set up your model such that it ballpark has a mean of 0 (e.g., by including it in `formula` too). -->

---

# Spatially varying effect of time

```{r plot-zeta}
qcs_grid$year_scaled <- (qcs_grid$year - mean(pcod$year)) / sd(pcod$year)
p <- predict(fit, qcs_grid)
p <- dplyr::filter(p, year == 2011) # an year
ggplot(p, aes(X, Y, fill = zeta_s)) + geom_raster() +
  scale_fill_gradient2()
```

<!-- See the vignette on [Fitting spatial trend models with sdmTMB](https://pbs-assess.github.io/sdmTMB/articles/spatial-trend-models.html) for more details. -->

---

# Spatially varying effect of NAO on Snowy Owl eruptions

```{r}
# code here, don't eval
```

---

# Spatially varying effect of NAO on Snowy Owl eruptions

```{r}
# plot here as png
```

---

# Tips and tricks

* Unlike `time_varying`, likely *do* want the same coefficient in the main `formula` effects
  * `spatial_varying` represents a random field with mean zero
* Predictor should be roughly mean zero and SD of 1 to help with estimation
* Current sdmTMB limited to one effect; an unlimited version is in development in a branch

See [vignette](https://pbs-assess.github.io/sdmTMB/articles/spatial-trend-models.html)
