---
title: "A simple spatial example"
author: "Sean Anderson"
output: html_document
---

```{r}
library(sdmTMB)
library(ggplot2)
library(dplyr)
```

Pick one year of data to model for this spatial example:

```{r}
d <- dplyr::filter(pcod, year == 2017)
```

We need to make a mesh, which has "knots", to it our model. These are a smaller number of locations than the data that the model will keep track of for computational efficiency. There are a number of ways of making meshes, which we can get into later.

```{r}
spde <- make_mesh(d, c("X", "Y"), cutoff = 5)
plot(spde)
```

We can fit a model similarly to how we would use glm(), or glmmTMB(), or lmer():

```{r}
m <- sdmTMB(
  data = d, 
  formula = density ~ s(log(depth), k = 3), # specify the 'k'!
  spde = spde, 
  silent = FALSE, # toggles whether you see progress while fitting
  family = tweedie(link = "log")
)
```

```{r}
print(m)
```

There is a built-in tidy() method, that is based on the broom package. This returns the model estimates in an easy-to-use data frame:

`?tidy.sdmTMB`

```{r}
tidy(m)
tidy(m, conf.int = TRUE)
tidy(m, effects = "ran_pars", conf.int = TRUE)
```

## Predictions at original data locations

Just like lm() or glm(), there is a predict() method. By default it will predict at the locations of the original data:

```{r}
predictions <- predict(m)
head(predictions)
```

The output is a data frame. It has the original data and a number of other columns:

- `est`: Estimate in link space
- `est_non_rf`: Estimate from everything that isn't a random field
- `est_rf`: Estimate from all random fields combined
- `omega_s`: Spatial (intercept) random field that is constant through time
- ignore: `zeta_s`: Spatial slope random field
- `epsilon_st`: Spatiotemporal (intercept) random fields (could be independent draws each year or AR1)

There is also a residuals method:

```{r}
predictions$resids <- residuals(m) # randomized quantile residuals
ggplot(predictions, aes(X, Y, col = resids)) + 
  scale_colour_gradient2() +
  geom_point() + facet_wrap(~year)
```

```{r}
hist(predictions$resids)
qqnorm(predictions$resids)
abline(a = 0, b = 1)
```

## Predictions onto new data

We can predict on new data. Here we will predict on a grid that is overlaid on the groundfish Queen Charlotte Sound synoptic survey domain:

```{r}
head(qcs_grid)
ggplot(qcs_grid, aes(X, Y, fill = -depth)) +
  geom_raster() +
  scale_fill_viridis_c() +
  coord_fixed()
```

```{r}
predictions <- predict(m, newdata = qcs_grid)
```

## Making maps of the output

Here's a short function for plotting our predictions:

```{r}
plot_map <- function(dat, column = est) {
  ggplot(dat, aes(X, Y, fill = {{column}})) + # {{}} uses tidyeval
    geom_raster() +
    coord_fixed()
}
```

These are the various spatial predictions are model has made:

```{r}
plot_map(predictions, exp(est)) +
  scale_fill_viridis_c(trans = "sqrt") +
  ggtitle("Prediction (fixed effects + all random effects)")
```

```{r}
plot_map(predictions, exp(est_non_rf)) +
  ggtitle("Prediction (fixed effects [depth])") +
  scale_fill_viridis_c(trans = "sqrt")
```

```{r}
plot_map(predictions, est_rf) +
  ggtitle("All random field estimates") +
  scale_fill_gradient2()
```

```{r}
plot_map(predictions, omega_s) +
  ggtitle("Spatial random effects only") +
  scale_fill_gradient2()
```

## Visualizing a conditional effect

We can visualize a conditional (sometimes called marginal) effect of our depth predictor by predicting across a range of depths with everything else held at a fixed value.

Make our prediction data frame:

```{r}
nd <- data.frame(
  depth = seq(min(d$depth), max(d$depth), length.out = 200))
head(nd)
```

Make the prediction. Note `re_form = NA` to avoid including the (random effect, 're') spatial component, which we don't want. It would be much slower if we did include it and we'd end up with a different height curve for each point in space.

```{r}
p <- predict(m, newdata = nd, se_fit = TRUE, re_form = NA)
ggplot(p, aes(
  x = depth, y = exp(est),
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  geom_ribbon(alpha = 0.4) +
  geom_line()
```
