---
title: Software setup for model-based estimates of abundance from survey data
---

1.  Install the following programs:

-   R 4.2.2 or other fairly recent version
-   Rtools 4.2 (or the version appropriate for an older version of R)
-   RStudio latest or recent version

It is important to be on these (or another very recent) version to ensure that INLA installs.

2.  In R, install the stable version of INLA (can take up to 10-15 minutes):

-   Install using instructions here: <https://www.r-inla.org/download-install>

Essentially you'll be running this:

```{r, eval=FALSE}
install.packages("INLA", repos = c(getOption("repos"), INLA = "https://inla.r-inla-download.org/R/stable"), dependencies = TRUE)
```

If you run into problems see the section ('INLA install problems') below.

Also install the mgcv package for fitting GAMs:

``` r
install.packages("mgcv", dependencies = TRUE)
```

3.  In R, install sdmTMB

If CRAN version is restored in time for the class session, sdmTMB can be installed from CRAN:

``` r
install.packages("sdmTMB", dependencies = TRUE)
```

If the CRAN version is not available, you will need to install the development version from GitHub.

Assuming you have a [C++ compiler](https://support.posit.co/hc/en-us/articles/200486498-Package-Development-Prerequisites) installed, install the development version:

```{r, eval=FALSE}
remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
```

-   If asked if you'd like to update any packages, say yes unless this may cause you problems in other projects.
-   If CRAN asks if you'd like to build a newer package from source, feel free to say no.
-   Sometimes some packages fail to install. If any packages fail (e.g. 'non-zero exist status), try installing them separately and then trying to install sdmTMB again.

E.g. if 'RcppParallel' and 'gap' failed to install:

```{r, eval=FALSE}
install.packages("RcppParallel")
install.packages("gap")
```

If you get the warning like this:

    TMB was built with Matrix version 1.5.3
    Current Matrix version is 1.5.1
    Please re-install 'TMB' from source using install.packages('TMB', type = 'source') or ask CRAN for a binary version of 'TMB' matching CRAN's 'Matrix' package

You can safely ignore it (if you really want to squash the warning, you can install TMB from source as described).

4.  Test to make sure everything is working by running this code:

```{r, eval=FALSE}
library(sdmTMB)
fit <- sdmTMB(
  density ~ s(depth),
  data = pcod_2011, 
  mesh = pcod_mesh_2011,
  family = tweedie(link = "log")
)
fit
```

If successful, you should see this:

    Spatial model fit by ML ['sdmTMB']
    Formula: density ~ s(depth)
    Mesh: pcod_mesh_2011
    Data: pcod_2011
    Family: tweedie(link = 'log')
     
                coef.est coef.se
    (Intercept)     2.16    0.34
    sdepth        -20.26   32.63

    Smooth terms:
               Std. Dev.
    sds(depth)     13.07

    Dispersion parameter: 13.68
    Tweedie p: 1.58
    Matern range: 16.84
    Spatial SD: 2.20
    ML criterion at convergence: 2937.789

    See ?tidy.sdmTMB to extract these values as a data frame.

Again, you can ignore any warnings about Matrix/TMB versions if you want.

## INLA install problems

First, make sure you are on the latest R (4.2.2) and Rtools (4.2).

Second, are you off any VPN? Sometimes that can block the INLA website.

If the install fails because the download times out, grab the URL to the .zip file and install it locally. E.g., as of today, this is the latest Windows version:

<https://inla.r-inla-download.org/R/stable/bin/windows/contrib/4.2/INLA_22.12.16.zip>

Download that to your Downloads folder, and install it with the following:

```{r, eval=FALSE}
user_folder <- Sys.getenv("USERPROFILE")
# adjust folder if needed:
zip_file <- file.path(user_folder, "Downloads/INLA_22.12.16.zip")

install.packages(zip_file, dependencies = TRUE, repos = NULL)
```
