---
title: "example"
author: "Sean Anderson"
output: html_document
---

```{r}
library(sdmTMB)
library(dplyr)
library(ggplot2)
```

UTMS!

```{r}
d <- dplyr::filter(pcod, year >= 2011) # subset for example speed
pcod_spde <- make_mesh(d, c("X", "Y"), cutoff = 20) # a coarse mesh for example speed
plot(pcod_spde)

# Tweedie:
m <- sdmTMB(density ~ 0 + depth_scaled + depth_scaled2 + as.factor(year),
  data = d, time = "year", spde = pcod_spde, family = tweedie(link = "log"))
print(m)
tidy(m, conf.int = TRUE)
tidy(m, effects = "ran_par", conf.int = TRUE)

# Bernoulli:
pcod_binom <- d
pcod_binom$present <- ifelse(pcod_binom$density > 0, 1L, 0L)
m_bin <- sdmTMB(present ~ 0 + as.factor(year) + depth_scaled + depth_scaled2,
  data = pcod_binom, time = "year", spde = pcod_spde,
  family = binomial(link = "logit"))
print(m_bin)

# Gaussian:
pcod_gaus <- subset(d, density > 0 & year >= 2013)
pcod_spde_gaus <- make_mesh(pcod_gaus, c("X", "Y"), cutoff = 30)
m_pos <- sdmTMB(log(density) ~ 0 + as.factor(year) + depth_scaled + depth_scaled2,
  data = pcod_gaus, time = "year", spde = pcod_spde_gaus)
print(m_pos)
```

With splines via mgcv.
Make sure to pre-specify an appropriate basis dimension (`k`) since
the smoothers are not penalized in the current implementation.
See ?mgcv::choose.k

```{r}
m_gam <- sdmTMB(log(density) ~ 0 + as.factor(year) + s(depth_scaled, k = 3),
  data = pcod_gaus, time = "year", spde = pcod_spde_gaus)
print(m_gam)
```

Brief aside on splines and basis functions:

```{r}
head(m_gam$tmb_data$X_ij)
basis1 <- m_gam$tmb_data$X_ij[,"s(depth_scaled).1"]
basis2 <- m_gam$tmb_data$X_ij[,"s(depth_scaled).2"]
plot(pcod_gaus$depth, basis1)
plot(pcod_gaus$depth, basis2)
```

Not penalized!

With IID random intercepts:

```{r}
set.seed(1)
x <- runif(500, -1, 1)
y <- runif(500, -1, 1)
loc <- data.frame(x = x, y = y)
spde <- make_mesh(loc, c("x", "y"), n_knots = 50, type = "kmeans")
s <- sdmTMB_sim(x = x, y = y, betas = 0, time = 1L,
  phi = 0.1, range = 1.4, sigma_O = 0.2, sigma_E = 0, mesh = spde)
s$g <- gl(50, 10)
iid_re_vals <- rnorm(50, 0, 0.3)
s$observed <- s$observed + iid_re_vals[s$g]
m <- sdmTMB(observed ~ 1 + (1 | g), spde = spde, data = s)
print(m)
tidy(m, "ran_pars", conf.int = TRUE) # see tau_G
theta <- as.list(m$sd_report, "Estimate")
plot(iid_re_vals, theta$RE)
```

We can fit a spatial model by omitting the time argument:

```{r}
m <- sdmTMB(
  density ~ depth_scaled + depth_scaled2, data = d,
  spde = pcod_spde, family = tweedie(link = "log"))
print(m)

# Spatial-trend example:
# m <- sdmTMB(density ~ depth_scaled, data = d,
#   spde = pcod_spde, family = tweedie(link = "log"),
#   spatial_trend = TRUE, time = "year")
# tidy(m, effects = "ran_par")

```

Time-varying effects of depth and depth squared:

```{r}
m <- sdmTMB(density ~ 0 + as.factor(year),
  time_varying = ~ 0 + depth_scaled + depth_scaled2,
  data = d, time = "year", spde = pcod_spde, family = tweedie(link = "log"))
print(m)
```

Pretty picture of that:

```{r}
nd <- expand.grid(
  year = unique(d$year),
  X = mean(d$X),
  Y = mean(d$Y),
  depth_scaled = seq(min(d$depth_scaled), max(d$depth_scaled), length.out = 100)
)
nd$depth_scaled2 <- nd$depth_scaled^2
p <- predict(m, newdata = nd, se_fit = TRUE, re_form = ~ 0)

ggplot(p, aes(x = depth_scaled, y = exp(est), group = year)) + 
  geom_ribbon(aes(ymax = exp(est + 2 * est_se), ymin = exp(est - 2 * est_se)), 
    alpha = 0.2) +
  geom_line(aes(colour = year)) +
  scale_colour_viridis_c()
```

# See the b_rw_t estimates; these are the time-varying (random walk) effects.
# These could be added to tidy.sdmTMB() eventually.
summary(m$sd_report)[1:19,]

# Linear breakpoint model on depth:
m_pos <- sdmTMB(log(density) ~ 0 + as.factor(year) +
    breakpt(depth_scaled) + depth_scaled2, data = pcod_gaus,
  time = "year", spde = pcod_spde_gaus)
print(m_pos)

# Linear covariate on log(sigma_epsilon):
# First we will center the years around their mean
# to help with convergence.
d$year_centered <- d$year - mean(d$year)
m <- sdmTMB(density ~ 0 + depth_scaled + depth_scaled2 + as.factor(year),
  data = d, time = "year", spde = pcod_spde, family = tweedie(link = "log"),
  epsilon_predictor = "year_centered")
print(m) # sigma_E varies with time now
# coefficient is not yet in tidy.sdmTMB:
as.list(m$sd_report, "Estimate", report = TRUE)$b_epsilon
as.list(m$sd_report, "Std. Error", report = TRUE)$b_epsilon

```

