---
title: sdmTMB setup
---

1. Install the following programs:

- R 4.2.2 or later
- Rtools 4.2 (or later to match your newer R version)
- RStudio latest version

It is important to be on this (or another very recent) version to ensure that INLA installs.

2. In R, install the stable version of INLA (can take 10-15 minutes especially if you're on a fresh setup):

- Install using information here: https://www.r-inla.org/download-install

Essentially you'll be running this:

```{r, eval=FALSE}
install.packages("INLA", repos = c(getOption("repos"), INLA = "https://inla.r-inla-download.org/R/stable"), dependencies = TRUE)
```

If you run into problems see the section ('INLA install problems') below.

3. In R, install sdmTMB from CRAN

```{r, eval=FALSE}
install.packages("sdmTMB", dependencies = TRUE)
```

- If asked if you'd like to update any packages, say yes.
- If CRAN asks if you'd like to build a newer package from source, feel free to say no.
- Sometimes some packages fail to install on DFO laptops. If any packages fail (e.g. 'non-zero exist status), try installing them separately and then trying to install sdmTMB again.

E.g. if 'RcppParallel' and 'gap' failed to install:

```{r, eval=FALSE}
install.packages("RcppParallel")
install.packages("gap")
```

Sometimes you may try restarting RStudio and/or restarting Windows to deal with folder permissions issues.

If you get the warning like this:

```
TMB was built with Matrix version 1.5.3
Current Matrix version is 1.5.1
Please re-install 'TMB' from source using install.packages('TMB', type = 'source') or ask CRAN for a binary version of 'TMB' matching CRAN's 'Matrix' package
```

You can safely ignore it (if you really want to squash the warning, you can install TMB from source as described).

4. Test to make sure everything is working:

```{r, eval=FALSE}
library(sdmTMB)
fit <- sdmTMB(
  density ~ s(depth),
  data = pcod_2011, 
  mesh = pcod_mesh_2011,
  family = tweedie(link = "log")
)
fit
```

If successful, you should see this:

```
Spatial model fit by ML ['sdmTMB']
Formula: density ~ s(depth)
Mesh: pcod_mesh_2011 (isotropic covariance)
Data: pcod_2011
Family: tweedie(link = 'log')
 
            coef.est coef.se
(Intercept)     2.16    0.34
sdepth         -1.94    3.13

Smooth terms:
           Std. Dev.
sds(depth)     13.07

Dispersion parameter: 13.68
Tweedie p: 1.58
Matérn range: 16.84
Spatial SD: 2.20
ML criterion at convergence: 2937.789

See ?tidy.sdmTMB to extract these values as a data frame.
```

Again, you can ignore any warnings about Matrix/TMB versions if you want.

## INLA install problems

We are very sorry if installing INLA gives you grief! Long term we'd like to remove this dependency.

First, make sure you are on the latest R (4.2.2) and Rtools (4.2).

If the install fails because the download times out, grab the URL to the .zip file and install it locally. E.g., as of today, this is the latest Windows version:

https://inla.r-inla-download.org/R/stable/bin/windows/contrib/4.2/INLA_22.12.16.zip 

Download that to your Downloads folder, and install it with the following:

```{r, eval=FALSE}
user_folder <- Sys.getenv("USERPROFILE")
# adjust folder if needed:
zip_file <- file.path(user_folder, "Downloads/INLA_22.12.16.zip")

install.packages(zip_file, dependencies = TRUE, repos = NULL)
```

## Installing the development version of sdmTMB

You can install development version of sdmTMB from GitHub. This requires a C++ compiler or the Rtools application on Windows.

```{r, eval=FALSE}
remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
remotes::install_github("pbs-assess/sdmTMBextra", dependencies = TRUE)
```
