---
title: "Building a prediction grid"
subtitle: "AMARE-MED workshop"
author: ""
institute: ""
date: "July 28--August 3 2023"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

<!-- Build with: xaringan::inf_mr() -->

```{r preamble, include=FALSE, cache=FALSE}
source("preamble.R")
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(dplyr)
library(sdmTMB)
library(ggplot2)
library(viridis)
```

# Why do we need a prediction grid?

* Generate some index of abundance

* Making pretty maps!

---

# Approach 1: at locations of data

```{r echo = TRUE, eval = FALSE}
pred <- predict(fit)
```

* Why? evaluating residuals

---

# Approach 2: mesh locations

* Predict to locations of mesh vertices

* These are where random effects are estimated 
  * spatial fields
  * spatiotemporal fields
  * spatially varying coefficients

* Why? Diagnostics / understanding mesh & uncertainties

See [https://github.com/pbs-assess/sdmTMB/discussions/230](https://github.com/pbs-assess/sdmTMB/discussions/230)
---

# Approach 3: regular grid

* see `qcs_grid` as example (4km grid)
* generally built from survey design, excludes land
* Why? Standardized index of abundance across survey domain

See `?get_index`
```{r echo=FALSE, eval=TRUE}
ggplot(qcs_grid, aes(X,Y, col = depth_scaled)) + 
  #geom_raster(alpha=0.5) + 
  geom_point()
```

---

# Approach 4: grid based on the locations of observations

* Things to consider: spatial grid cell size?
* How many time steps should have observations?
* Why? Domain grid not available, applicable
  * e.g. rock / reef associated fishes

Code: [https://github.com/pbs-assess/sdmTMB-teaching/blob/main/imr-2023/exercises/coastal-survey-ex/coastal-survey-index-south.R#L205](https://github.com/pbs-assess/sdmTMB-teaching/blob/main/imr-2023/exercises/coastal-survey-ex/coastal-survey-index-south.R#L205)

---

# Tips and tricks

* Any prediction grid or dataframe will need to have values for all covariates
  * Excamples: year, depth, temperature
  * Random effects = exception, can be NA
  
