---
title: "Introduction to spatial models with sdmTMB"
subtitle: "IMR sdmTMB workshop"
author: ""
institute: ""
date: "May 23--25 2023"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

<!-- Build with: xaringan::inf_mr() -->

```{r preamble, include=FALSE, cache=FALSE}
source("preamble.R")
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(dplyr)
library(sdmTMB)
library(ggplot2)
```

# The Pacific Cod dataset

DFO trawl survey data from Queen Charlotte Sound off BC

.small[
```{r, echo=TRUE}
library(dplyr)
library(ggplot2)
library(sdmTMB)
pcod %>%
  select(year, X, Y, depth, density, present) %>% 
  head()
```
]

---

# Building a mesh

.small[
```{r, echo=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
plot(mesh)
```
]

---

# Adding spatial random fields

```{r pcod-null, echo=TRUE}
fit <- sdmTMB(
  present ~ 1, # intercept only
  data = pcod,
  mesh = mesh,
  family = binomial(link = "logit"),
  spatial = "on"   #<<
)
```

---

# Should we include depth?

* Predictions here in *logit* space  

```{r plotdepth, echo=FALSE}
pred_df <- pcod
pred_df$pred <- predict(fit)$est

ggplot(pred_df, aes(depth, pred)) + 
  geom_point(size=0.3, alpha=0.5) + 
  geom_smooth(se = FALSE) + 
  xlab("pcod$depth") + 
  ylab("predict(fit)$est")
```

---

# Add depth as a quadratic effect

```{r pcod-spatial-rf-depth, echo=TRUE}
fit <- sdmTMB(
  present ~ poly(log(depth), 2), #<<
  data = pcod,
  mesh = mesh,
  family = binomial(link = "logit"),
  spatial = "on"   #<<
)
```

---

# Checking convergence  

* Learning curve of TMB can be steep  
* Models can be very complex (hard to diagnose issues)
* `sanity()` function tries to help

.small[
```{r checksanity, eval=TRUE, message=TRUE, warning=TRUE, echo=TRUE}
sanity(fit)
```
]

---

# Inspecting the model output

.small[
```{r, echo=TRUE, highlight.output=c(1)}
fit
```
]

---

# Inspecting the model output

.small[
```{r, echo=TRUE, highlight.output=2:5}
fit
```
]

---

# Inspecting the model output

.small[
```{r, echo=TRUE, highlight.output=7:10}
fit
```
]
---

# Inspecting the model output

.small[
```{r, echo=TRUE, highlight.output=12:14}
fit
```
]

---

# Extracting parameters in a data frame

```{r, echo=TRUE}
# Main effects:
tidy(fit)

# Variance-related terms:
tidy(fit, "ran_pars", conf.int = TRUE)
```

---

# Making predictions

.small[
```{r, echo=TRUE}
p <- predict(fit) #<<
select(p, X, Y, est:omega_s)
```
]

---

# Making predictions

* `est`: Overall estimate **in link space** (logit here)
* `est_non_rf`: Estimate of non-random-field components
* `est_rf`: Estimate of random-field components
* `omega_s`: Spatial random field

---

# Making predictions on new data

.small[
* `qcs_grid`: a 2x2 km grid extending over the full survey domain
]
.small[
```{r, echo=TRUE, out.width="500px", fig.width=5}
data("qcs_grid")
nd <- filter(qcs_grid)
ggplot(nd, aes(X, Y, fill = depth)) + 
  geom_raster() +
  coord_fixed()
```
]

---

# Making predictions on new data


```{r, echo=TRUE}
head(select(nd, X, Y, depth))

p <- predict(fit, newdata = nd) #<<
```

---

# Plotting predictions on new data

.small[
```{r, echo=TRUE, fig.width=6, out.width="600px"}
ggplot(p, aes(X, Y, fill = plogis(est))) +
  geom_raster() +
  scale_fill_viridis_c() +
  coord_fixed()
```
]

---

# Plotting main effect contributions

.small[
```{r, echo=TRUE, fig.width=6, out.width="600px"}
ggplot(p, aes(X, Y, fill = plogis(est_non_rf))) +
  geom_raster() +
  scale_fill_viridis_c() +
  coord_fixed()
```
]

---

# Plotting spatial random effects

.small[
```{r, echo=TRUE, fig.width=6, out.width="600px"}
ggplot(p, aes(X, Y, fill = omega_s)) +
  geom_raster() +
  scale_fill_gradient2() +
  coord_fixed()
```
]

---

# Plotting depth effect

.small[
```{r pcod-sp-depth-pred, echo=TRUE}
nd <- data.frame(depth = seq(30, 500, length.out = 300))
p <- predict(
  fit, 
  newdata = nd, 
  se_fit = TRUE, #<<
  re_form = ~ 0 #<<
)
p$lwr <- plogis(p$est - 1.96 * p$est_se)
p$upr <- plogis(p$est + 1.96 * p$est_se)
```

```{r pcod-sp-depth-pred-plot, fig.width=5, echo=FALSE, out.width='500px', fig.asp=0.5}
ggplot(p, aes((nd$depth), plogis(est), ymin = lwr, ymax = upr)) +
  geom_ribbon(alpha = 0.2) +
  geom_line() + xlab("Depth (m)") + ylab("") +
  coord_cartesian(ylim = c(0, 1), expand = FALSE)
```
]

---

# Plotting depth effect with visreg

.small[
```{r pcod-sp-depth-visreg, echo=TRUE}
visreg::visreg(fit, xvar = "depth", gg = TRUE)
```
]

---

# Plotting depth effect with visreg

.small[
```{r pcod-sp-depth-visreg-resp, echo=TRUE}
visreg::visreg(fit, xvar = "depth", scale = "response")
```
]

---

# Plotting depth effect with ggeffects

.small[
```{r pcod-sp-depth-ggeffects, echo=TRUE}
g <- ggeffects::ggeffect(fit, "depth [10:500 by=2]")
plot(g)
```
]
