---
title: "Area-weighted index standardization"
subtitle: "DFO TESA sdmTMB workshop"
author: ""
institute: ""
date: "January 16--19 2023"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

<!-- Build with: xaringan::inf_mr() -->

```{r preamble, include=FALSE, cache=FALSE}
source(here::here("dfo-tesa-2023/preamble.R"))
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(dplyr)
library(sdmTMB)
library(ggplot2)
```

# Why model-based index standardization?

* Design-based methods (stratified means) are straightforward but assume perfect survey implementation

--

* Model-based approaches can account for (some) aspects of imperfect implementation

--

* Model-based approaches can leverage spatial correlation to improve precision

--

* Model-based approaches can integrate additional data sources, multiple gear types, accounting for gear calibration, etc.

---

# Design- vs. model-based estimators

--

* Design-based: fish are homogenously distributed within strata; noise comes from sampling

--

* Model-based: fish are distributed according to a mathematical model; noise from uncertainty in fish distribution and sampling

---

# Calculating an area-weighted population index

1. Fit a sensible model: .blue[`sdmTMB()`]

2. Predict on a grid covering the survey domain: .blue[`predict(..., newdata = ...)`]

3. Sum up the density multiplied by cell area to calculate total abundance or biomass: .blue[`get_index()`]

---

# Fit a sensible model

```{r pcod-st-fit, echo=TRUE, eval=TRUE, results='hide'}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
fit <- sdmTMB(
  density ~ 0 + as.factor(year), #<<
  data = pcod,
  mesh = mesh,
  family = tweedie(link = "log"),
  spatial = "on",
  time = "year",
  spatiotemporal = "iid", #<<
  silent = FALSE # show progress!
)
```

---

# Predict over the survey domain

.small[
```{r pcod-st-index1, echo=TRUE, message=FALSE, warning=FALSE}
pred <- predict(
  fit, newdata = qcs_grid, return_tmb_object = TRUE
)
select(pred$data, year, X, Y, est) |> 
  head()
```
]

---

# Predict over the survey domain

```{r pcod-st-plot-est2, echo=FALSE, fig.width=7}
.p <- pred$data
max_est <- quantile(.p$est, 0.998)
.p <- mutate(.p, est_trim = if_else(est > max_est, max_est, est))
ggplot(.p, aes(X, Y, fill = exp(est_trim))) +
  geom_raster() +
  scale_fill_viridis_c(trans = "sqrt") +
  facet_wrap(~year) +
  coord_fixed() +
  labs(fill = "Biomass density\n(kg/km^2)")
```

---

# Sum up the density multiplied by cell area

(*and calculate standard errors)

```{r pcod-index, echo=TRUE}
index <- get_index(pred, area = 4, bias_correct = TRUE)
head(index)
```

---

# The resulting standardized index

.xsmall[
```{r pcod-st-index-plot, echo=TRUE, fig.width=6, fig.asp=0.5, out.width="600px"}
ggplot(index, aes(year, est)) + geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
  xlab('Year') + ylab('Biomass estimate (kg)')
```
]

---

# Remember...

There is *one* design-based index.

--

A good model-based estimated depends on the modeller constructing a reasonable model!

--

There are *many* possible model-based indexes. Thankfully, results are often (but not always!) qualitatively similar across configurations.


.tiny[
Commander, C.J.C, L.A.K. Barnett, E.J. Ward, S.C. Anderson, T.E. Essington. 2022. The shadow model: how and why small choices in spatially explicit species distribution models affect predictions. PeerJ. 10: e12783. <https://doi.org/10.7717/peerj.12783>
]
