---
title: "Simulating and fitting spatial random fields"
format: html
editor: visual
execute: 
  echo: true
  eval: true
---

# Goals:

-   Understand the spatial range.
-   Become familiar with spatial random fields through simulation.
-   Become familiar with spatiotemporal random fields and how they can combine with spatial random fields.
-   Understand how correlation in spatiotemporal random fields looks.

```{r}
#| echo=FALSE
library(sdmTMB)
library(ggplot2)
options(ggplot2.continuous.colour = "viridis")
options(ggplot2.continuous.fill = "viridis")
theme_set(theme_light())
```

# Understanding the spatial range

The spatial range is the distance at which two data points are effectively independent. Technically, this is the distance at which correlation declines to approximately 0.13.

We can visualize the Matérn correlation function for various range values:

```{r}
range <- 0.5 #< try changing this

kappa <- sqrt(8) / range
distance <- seq(0.001, 1, length.out = 100)
correlation <- kappa * distance * besselK(kappa * distance, nu = 1)
plot(distance, correlation, type = "l", ylim = c(0, 1))
abline(h = 0.13, lty = 2) # 0.13 correlation
abline(v = range, lty = 2)
```

Here we have a unitless distance from 0 to 1. In real situations these would be in something like m or km and so the range would be units of m or km.

### Exercise:

1.  Try running the previous chunk over and over with various `range` values. Note how the correlation function changes.

# Spatial simulation

Next we will simulate some spatial random fields to understand what they look like and how their parameters control them.

First, this chunk creates a set of X and Y spatial coordinates and constructs a mesh using a minimum triangle edge length cutoff distance of 0.05. We need this for the internal random field calculations with the SPDE.

```{r}
# A grid of X and Y values:
predictor_dat <- expand.grid(
  X = seq(0, 1, length.out = 100), Y = seq(0, 1, length.out = 100)
)
mesh <- make_mesh(predictor_dat, c("X", "Y"), cutoff = 0.05)
plot(mesh)
```

The basic function for simulating random fields in sdmTMB is `sdmTMB_simulate()`.

-   The formula here contains only an intercept but could include covariates.

-   The family is Gaussian to specify that we simulate a normally distributed response.

-   The range (see above) is 0.6.

-   The two important standard deviation parameters are `sigma_O` (controlling the marginal standard deviation of `omega`, the spatial random field) and `phi` (controlling the observation error standard deviation).

-   We are plotting `observed` (observed values: the expected value plus observation error).

```{r}
sim_dat <- sdmTMB_simulate(
  formula = ~ 1,
  data = predictor_dat,
  mesh = mesh,
  family = gaussian(link = "identity"),
  range = 0.6, #< try changing this (spatial range)
  sigma_O = 0.2, #< try changing this (spatial standard deviation)
  phi = 0.001, #< try changing this observation error standard deviation
  B = 0 # intercept
)
# head(sim_dat)

ggplot(sim_dat, aes(X, Y, fill = observed)) +
  geom_raster()
```

### Exercise:

1.  Try running the previous chunk over and over. Try changing the `range` and `sigma_O` values. What do you observe?
2.  Try increasing the observation error standard deviation `phi` and re-running it many times. Try making it equal to `sigma_O` and twice its size. What happens when observation error is much larger than `sigma_O`?

Note that the colour scale may be changing between runs!

# Spatiotemporal simulation

```{r}
predictor_dat2 <- expand.grid(
  X = seq(0, 1, length.out = 100), Y = seq(0, 1, length.out = 100), year = 1:6
)
mesh2 <- make_mesh(predictor_dat2, c("X", "Y"), cutoff = 0.1)
```

```{r}
sim_dat2 <- sdmTMB_simulate(
  formula = ~ 1,
  data = predictor_dat2,
  time = "year",
  mesh = mesh2,
  family = gaussian(),
  range = 0.5, #< spatial/spatiotemporal range
  sigma_O = 0.2, #< spatial SD
  sigma_E = 0.1, #< spatiotemporal SD
  phi = 0.001,
  B = 0 # intercept
)
# head(sim_dat2)
```

```{r spatial-fields}
ggplot(sim_dat2, aes(X, Y, fill = omega_s)) +
  geom_raster() +
  facet_wrap(~year) +
  scale_fill_gradient2()
```

```{r spatiotemporal-fields}
ggplot(sim_dat2, aes(X, Y, fill = epsilon_st)) +
  geom_raster() +
  facet_wrap(~year) +
  scale_fill_gradient2()
```

```{r simulated-observations}
ggplot(sim_dat2, aes(X, Y, fill = observed)) +
  geom_raster() +
  facet_wrap(~year) +
  scale_fill_gradient2()
```

### Exercise:

1.  Try running the previous 4 chunks over and over and observe the various components.
2.  What do you notice about `omega_s` (spatial random field) across the time slices? Conversely, what do you notice about `epsilon_st` (spatiotemporal random fields) across time?
3.  Try increasing `sigma_E` (spatiotemporal SD) to be larger than `sigma_O` (spatial SD). How do the observations change?

# Correlated spatiotemporal fields

The above spatiotemporal fields were left as their default: independent across years (`spatiotemporal = "iid"`). They can also be correlated. Here we will simulate and plot spatiotemporal fields with various levels of correlation.

```{r}
sim_dat3 <- sdmTMB_simulate(
  formula = ~ 1,
  data = predictor_dat2,
  time = "year",
  mesh = mesh2,
  family = gaussian(),
  range = 0.5,
  sigma_O = 0,
  sigma_E = 0.1,
  spatiotemporal = "ar1", #< field type
  rho = 0.6, #< AR(1) field correlation
  phi = 0.001,
  B = 0 # intercept
)
ggplot(sim_dat3, aes(X, Y, fill = epsilon_st)) +
  geom_raster() +
  facet_wrap(~year) +
  scale_fill_gradient2()
```

### Exercise:

1.  In the previous chunk, try setting the AR(1) random field correlation (`rho`) to values ranging from 0 to near 1 (e.g., 0.99). What do you observe?
2.  When might you observe a process that resembles the above?
3.  What happens when `rho` is \< 0? Is `rho` more likely to be \> or \< than 0 in ecological systems? Can you think of a situation where `rho < 0`?

# Fitting a geostatistical spatial model

Let's fit a basic spatial model to some simulated data. We'll simulate on a grid and then down-sample from that grid. (We'll do it this way so we can plot a smooth known random field in the end for comparison.)

```{r}
predictor_dat <- expand.grid(
  X = seq(0, 1, length.out = 100), 
  Y = seq(0, 1, length.out = 100)
)
set.seed(19201) # for consistency
predictor_dat$covariate <- rnorm(n = nrow(predictor_dat))
mesh <- make_mesh(predictor_dat, c("X", "Y"), cutoff = 0.05)
plot(mesh)

sim_dat4 <- sdmTMB_simulate(
  formula = ~ 1 + covariate,
  data = predictor_dat,
  mesh = mesh,
  family = gaussian(link = "identity"),
  range = 0.5,
  sigma_O = 0.4,
  phi = 0.3,
  B = c(0.4, 0.2), # intercept and covariate coefficient
  seed = 98259 # for consistency
)
head(sim_dat4)

ggplot(sim_dat4, aes(X, Y, fill = omega_s)) +
  geom_raster()

ggplot(sim_dat4, aes(X, Y, fill = mu)) +
  geom_raster()

ggplot(sim_dat4, aes(X, Y, fill = observed)) +
  geom_raster()
```

### Exercise:

1.  What did we just simulate above? What is the one addition from the first spatial simulation?
2.  What do the last 3 plots represent? Why do they get increasingly noisy?

Now we'll down-sample to simulate sampling/observing only 300 locations:

```{r}
set.seed(1098)
to_sample <- sample(seq_len(nrow(sim_dat4)), size = 300)
obs_dat <- sim_dat4[to_sample, ]
ggplot(obs_dat, aes(X, Y, colour = observed)) +
  geom_point()
```

Start by fitting a regular GLM:

```{r}
fit_glm <- glm(observed ~ 1 + covariate, family = gaussian(), data = obs_dat)
summary(fit_glm)
```

We can calculate and plot the residuals spatially:

```{r}
obs_dat$resid <- residuals(fit_glm)
ggplot(obs_dat, aes(X, Y, colour = resid)) +
  geom_point() +
  scale_colour_gradient2()
```

And plot predictions spatially on the full grid:

```{r}
sim_dat4$pred_glm <- predict(fit_glm, newdata = sim_dat4)

lims <- range(c(sim_dat4$mu, sim_dat4$pred_glm))

g1 <- ggplot(sim_dat4, aes(X, Y, fill = mu)) +
  geom_raster() + ggtitle("Truth") +
  scale_fill_viridis_c(limits = lims)
g2 <- ggplot(sim_dat4, aes(X, Y, fill = pred_glm)) +
  geom_raster() + ggtitle("Estimated") +
  scale_fill_viridis_c(limits = lims)
cowplot::plot_grid(g1, g2, nrow = 2)
```

### Exercise:

1.  Do you see spatial correlation in the residuals?
2.  What are at least 2 problems this creates?
3.  How good do those predictions look compared to the known truth?
4.  In the context of this workshop, what is a solution to this?

First, let's convince ourselves that we can fit the exact same model with sdmTMB. All we have to do is turn off the spatial random fields:

```{r}
fit_glm2 <- sdmTMB(observed ~ 1 + covariate, family = gaussian(), 
  data = obs_dat, spatial = "off") # default is spatial = "on"
summary(fit_glm2)
```

### Exercise:

1.  Check if `fit_glm` and `fit_glm2` match and represent the same model.

Now, try fitting a GLM with Gaussian random fields:

```{r}
fitting_mesh <- make_mesh(obs_dat, c("X", "Y"), cutoff = 0.05)
plot(fitting_mesh)
fit_sdmTMB <- sdmTMB(
  observed ~ 1 + covariate, 
  family = gaussian(), 
  data = obs_dat, 
  spatial = "on",
  mesh = fitting_mesh
)
sanity(fit_sdmTMB)
summary(fit_sdmTMB)
tidy(fit_sdmTMB, conf.int = TRUE)
tidy(fit_sdmTMB, "ran_pars", conf.int = TRUE)
```

```{r}
summary(fit_glm2)
summary(fit_sdmTMB)
```

### Exercise:

1.  How did the dispersion parameter (Gaussian SD) differ between `fit_glm2` and `fit_sdmTMB`? Why?
2.  How close are the estimated values to the truth? Is that about what you expected?

Now let's calculate and plot residuals for the random field model:

```{r}
obs_dat$resid_sdmTMB <- residuals(fit_sdmTMB)
ggplot(obs_dat, aes(X, Y, colour = resid_sdmTMB)) +
  geom_point() +
  scale_colour_gradient2()
```

And make predictions on the full grid:

```{r}
p <- predict(fit_sdmTMB, newdata = sim_dat4)
```

```{r}
lims <- range(c(sim_dat4$mu, p$est))
g1 <- ggplot(sim_dat4, aes(X, Y, fill = mu)) +
  geom_raster() + ggtitle("True mean") +
  scale_fill_viridis_c(limits = lims)

g2 <- ggplot(p, aes(X, Y, fill = est)) +
  geom_raster() + ggtitle("Estimated value") +
  scale_fill_viridis_c(limits = lims)

cowplot::plot_grid(g1, g2, nrow = 2)
```

```{r}
lims <- range(c(sim_dat4$omega_s, p$omega_s))
g1 <- ggplot(sim_dat4, aes(X, Y, fill = omega_s)) +
  geom_raster() + ggtitle("True random field") +
  scale_fill_gradient2(limits = lims)

g2 <- ggplot(p, aes(X, Y, fill = omega_s)) +
  geom_raster() + ggtitle("Estimated random field") +
  scale_fill_gradient2(limits = lims)
cowplot::plot_grid(g1, g2, nrow = 2)
```

Predicted vs. true for the two models:

```{r}
par(mfrow = c(2, 1))
plot(sim_dat4$mu, sim_dat4$pred_glm, main = "GLM", 
  xlab = "True", ylab = "Predicted")
plot(sim_dat4$mu, p$est, main = "GLM with random fields", 
  xlab = "True", ylab = "Predicted")
```

### Exercise:

1.  How do the residuals look now? Why?
2.  How do the predictions look compared to before? Why?
3.  How do the random fields look? Is this as good as you expected? What would make these easier/harder to estimate?
