---
title: "Fitting spatial/spatiotemporal models and problem solving"
format: html
editor: visual
execute:
  echo: true
  eval: true
---

# Goals:

-   Practice fitting spatial and spatiotemporal models with less guidance.
-   See an example of using an 'offset'.
-   Practice calculating an area-weighted biomass index with less guidance.
-   Gain some problem solving skills by fixing 3 models with estimation problems or other errors you may encounter.

```{r, message=FALSE, warning=FALSE}
library(sdmTMB)
library(dplyr)
library(ggplot2)
options(ggplot2.continuous.colour = "viridis")
options(ggplot2.continuous.fill = "viridis")
theme_set(theme_light())
```

We will work with lingcod data from the Queen Charlotte Sound synoptic survey:

# Data

```{r}
d <- readRDS(here::here("dfo-tesa-2023/data/lingcod-qcs.rds"))
d$area_swept <- d$area_swept / 10000
glimpse(d)
```

Add UTM columns to the dataset. Use `utm_crs = 3156` to match the UTM zone 9 used in the grid.

```{r}
d <- add_utm_columns(d, units = "km", utm_crs = 3156) # excercise
```

Create a mesh with a `cutoff` value of `10` km.

# Mesh

```{r}
mesh <- make_mesh(d, c("X", "Y"), cutoff = 10) # excercise
plot(mesh)
```

We will run the following code to turn the `qcs_grid` grid built into the package into a grid with values for every year in the lingcod dataset.

# Grid

```{r}
grid <- filter(qcs_grid, year == min(qcs_grid$year)) |>
  select(-year) |>
  select(-depth_scaled, -depth_scaled2) |>
  rename(depth_m = depth)
grid <- purrr::map_dfr(unique(d$year), ~ tibble(grid, year = .x))
```

# Spatial model

This is how an 'offset' works:

-   log(catch_weight/area_swept) = b0 + b1 \* x
-   log(catch_weight) - log(area_swept) = b0 + b1 \* x
-   log(catch_weight) = b0 + b1 \* x + log(area_swept)
-   log(catch_weight) = b0 + b1 \* x + offset

Note the offset is the **log** of the effort variable.

Here, we could fit our model to catch weight with an area swept offset or to density. In other cases (e.g., count data), an offset lets us use a count distribution with an effort measure.

Fit a spatial model with a delta-Gamma family:

```{r}
fit_spatial <- sdmTMB(
  catch_weight ~ 0 + as.factor(year),
  spatial = "on", # answer
# spatial = , # exercise
  offset = log(d$area_swept),
  family = delta_gamma(), # answer
# family =  , # exercise
  mesh = mesh, # answer
# mesh =  , # exercise
  data = d, # answer
  # silent = FALSE,
  control = sdmTMBcontrol(newton_loops = 1) # helps with convergence here
)
```

```{r}
sanity(fit_spatial)
fit_spatial
```

Calculate residuals and assign them to a new column `resid_sp` in `d`:

```{r}
d$resid_sp <- residuals(fit_spatial) # answer
ggplot(d, aes(X, Y, colour = resid_sp)) +
  geom_point() +
  facet_wrap(~year) +
  scale_colour_gradient2()
```

### Exercise:

1.  Did that fit reasonably well such that we can try adding the complexity of spatiotemporal fields?
2.  In addition to the fits, do the spatial residuals in any year show evidence of persistent spatial autocorrelation?

# Spatiotemporal model

Add IID spatiotemporal fields:

```{r}
fit <- sdmTMB(
  catch_weight ~ 0 + as.factor(year),
  time = "year", # answer
# time = , # exercise
  spatial = "on", # answer
# spatial = , # exercise
  spatiotemporal = "iid", # answer
# spatiotemporal = , # exercise
  offset = log(d$area_swept),
  family = delta_gamma(), # answer
# family = , # exercise
  mesh = mesh, # answer
# mesh = , # exercise
  data = d,
  # silent = FALSE,
  control = sdmTMBcontrol(newton_loops = 1) # helps with convergence here
)
```

```{r}
sanity(fit)
```

```{r}
fit
```

```{r}
tidy(fit, model = 1, conf.int = TRUE)
tidy(fit, model = 2, conf.int = TRUE)

tidy(fit, "ran_pars", model = 1, conf.int = TRUE)
tidy(fit, "ran_pars", model = 2, conf.int = TRUE)
```

### Exercise:

1.  Is the range bigger or smaller for the binomial component than the Gamma component?
2.  Is the spatial or spatiotemporal variance larger?
3.  Does that make sense? What does this mean biologically?

Calculate randomized quantile residuals into an object `resid`:

```{r}
set.seed(1)
resid <- residuals(fit) # answer
#resid <-  # exercise
```

How do the basic randomized quantile residuals look for the overall model?

```{r}
qqnorm(resid)
qqline(resid)
```

Predict on the newdata `grid` and set `return_tmb_object = TRUE`. Save the output to `pred`:

```{r}
pred <- predict(fit, newdata = grid, return_tmb_object = TRUE) # answer
#pred <-  # exercise
```

Because we set `return_tmb_object = TRUE`, the prediction data frame is in the list element `data`.

```{r}
glimpse(pred$data)
```

### Exercise:

1.  What do the `1` and `2` columns represent?
2.  How do we turn these into response space?

## Plotting the model components spatially

```{r}
ggplot(pred$data, aes(X, Y, fill = plogis(est_non_rf1))) +
  geom_raster() +
  facet_wrap(~year) +
  coord_fixed() +
  ggtitle("Fixed effect components 1")
```

```{r}
ggplot(pred$data, aes(X, Y, fill = exp(est_non_rf2))) +
  geom_raster() +
  facet_wrap(~year) +
  coord_fixed() +
  ggtitle("Fixed effect components 2")
```

```{r}
ggplot(pred$data, aes(X, Y, fill = omega_s2)) +
  geom_raster() +
  scale_fill_gradient2() +
  coord_fixed() +
  ggtitle("Spatial random field 2")
```

```{r}
ggplot(pred$data, aes(X, Y, fill = epsilon_st1)) +
  geom_raster() +
  scale_fill_gradient2() +
  facet_wrap(~year) +
  coord_fixed() +
  ggtitle("Spatiotemporal random fields 1")
```

```{r}
ggplot(pred$data, aes(X, Y, fill = epsilon_st2)) +
  geom_raster() +
  scale_fill_gradient2() +
  facet_wrap(~year) +
  coord_fixed() +
  ggtitle("Spatiotemporal random fields 2")
```

```{r}
ggplot(pred$data, aes(X, Y, fill = plogis(est1))) +
  geom_raster() +
  facet_wrap(~year) +
  coord_fixed() +
  ggtitle("Prediction 1")
```

```{r}
ggplot(pred$data, aes(X, Y, fill = exp(est2))) +
  geom_raster() +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10") +
  coord_fixed() +
  ggtitle("Prediction 2")
```

```{r}
ggplot(pred$data, aes(X, Y, fill = plogis(est1) * exp(est2))) +
  geom_raster() +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10", option = "C") +
  coord_fixed() +
  ggtitle("Combined prediction")
```

# Calculating an area-weighted biomass index

Use `get_index()` to calculate an area-weighted biomass index. Feel free to set `bias_correct = FALSE` to be a bit faster. Save the output to `ind`.

```{r}
ind <- get_index(pred, area = 4, bias_correct = FALSE) # answer
#ind <-  # exercise
```

```{r}
ggplot(ind, aes(year, y = est, ymin = lwr, ymax = upr)) +
  geom_pointrange() +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05)))
```

# Problem solving

The following are 3 models where each has some estimation problem or creates an error. Try to fix the problems.

## Problem 1:

```{r}
glimpse(pcod) # built in
```

```{r}
ggplot(pcod, aes(X, Y, size = density)) +
  geom_point(shape = 21)
```

```{r}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 40)
plot(mesh)
```

```{r}
fit_pcod <- sdmTMB(
  density ~ poly(depth, 2),
  spatial = "on",
  family = tweedie(link = "log"),
  mesh = mesh,
  data = pcod
)
```

```{r}
sanity(fit_pcod)
fit_pcod
```

### Exercise:

1.  What is the problem with the above?

-   Hint: look at `sanity(fit_pcod)`
-   Hint: look at the range estimate with `print(fit_pcod)`
-   Hint: look at the mesh `cutoff` value

2.  Try fixing it below:

```{r}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10) # answer
#mesh <- # exercise
plot(mesh)
```

```{r}
fit_pcod2 <- sdmTMB(
  density ~ poly(depth, 2),
  spatial = "on",
  family = tweedie(link = "log"),
  mesh = mesh,
  data = pcod
)
```

```{r}
sanity(fit_pcod2)
fit_pcod2
```

## Problem 2

The following is some yelloweye rockfish data from the hard bottom longline survey on the inside waters of Vancouver Island.

```{r}
d <- readRDS(here::here("dfo-tesa-2023/data/hbll-inside-yelloweye.rds"))
d_north <- dplyr::filter(d, survey == "HBLL INS N")
d_north <- add_utm_columns(d_north)
```

```{r}
ggplot(d_north, aes(X, Y, size = catch_count)) +
  facet_wrap(~year) +
  geom_point(shape = 21)
```

```{r}
mesh_north <- make_mesh(d_north, xy_cols = c("X", "Y"), cutoff = 5)
plot(mesh_north)
fit_ye <- sdmTMB(
  catch_count ~ 0 + as.factor(year),
  offset = log(d_north$hook_count),
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  family = nbinom2(link = "log"),
  # silent = FALSE,
  mesh = mesh_north,
  data = d_north
)
```

```{r}
sanity(fit_ye)
```

### Exercise:

1.  What is the problem with the above?
2.  What are some ways you could attempt to fix this?
3.  What is a simpler random field structure you could try? Try that below with `fit_ye1`.
4.  What is a simpler family you could try? Try that below with `fit_ye2`.

```{r}
fit_ye1 <- update(fit_ye, spatiotemporal = "off") # answer
#fit_ye1 <- update(fit_ye, spatiotemporal = ) # exercise
sanity(fit_ye1)
fit_ye1
```

```{r}
fit_ye2 <- update(fit_ye, family = poisson(link = "log")) # answer
#fit_ye2 <- update(fit_ye, family = ) # exercise
sanity(fit_ye2)
fit_ye2
```

## Problem 3

```{r}
ggplot(pcod, aes(X, Y, colour = present)) +
  geom_point()
```

```{r}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
```

```{r}
dat <- filter(pcod, year == 2017)
```

```{r, eval=FALSE}
fit_pcod <- sdmTMB(
  present ~ poly(depth, 2),
  spatial = "on",
  family = binomial(),
  mesh = mesh,
  data = dat
)
```

### Exercise:

1.  What is causing the error message?
2.  Try to fix it below.

Hint: When was the mesh created relative to any filtering of the data? Why is this a problem?

```{r}
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 10) # answer
# exercise
fit_pcod <- sdmTMB(
  present ~ poly(depth, 2),
  spatial = "on",
  family = binomial(),
  mesh = mesh,
  data = dat
)
fit_pcod
```

Bonus: try plotting the depth effect using the `ggeffect::ggeffect()` function.

```{r}
ggeffects::ggeffect(fit_pcod, "depth [0:600 by=2]") |> plot() # answer
## exercise
```
