---
title: "Example: Spatiotemporal modeling of Pacific cod"
subtitle: "DFO TESA sdmTMB workshop"
author: ""
institute: ""
date: "January 16--19 2023"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

<!-- Build with: xaringan::inf_mr() -->

```{r preamble, include=FALSE, cache=FALSE}
source(here::here("dfo-tesa-2023/preamble.R"))
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(dplyr)
library(sdmTMB)
library(ggplot2)

mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
```

# Extending to a spatiotemporal model

Reminder, the Pacific Cod dataset:

.small[
```{r, echo=TRUE}
pcod %>% 
  select(year, X, Y, depth, density, present) %>% 
  head()

mesh <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 10)
```
]

---

# Our previous spatial model fit

```{r pcod-spatial-rf-reminder, echo=TRUE, eval=FALSE}
fit <- sdmTMB(
  present ~ poly(log(depth), 2),
  data = pcod,
  mesh = mesh,
  family = binomial(link = "logit"),
  spatial = "on"
)
```

---

# Switching to density

```{r pcod-spatial-rf-density, echo=TRUE, eval=FALSE}
fit <- sdmTMB(
  density ~ poly(log(depth), 2), #<<
  data = pcod,
  mesh = mesh,
  family = tweedie(link = "log"), #<<
  spatial = "on"
)
```

---

# Adding spatiotemporal fields

```{r pcod-st-fit, echo=TRUE, eval=TRUE, results='hide'}
fit <- sdmTMB(
  density ~ s(depth, k = 5) + 0 + as.factor(year), #<<
  data = pcod,
  mesh = mesh,
  family = tweedie(link = "log"),
  spatial = "on",
  time = "year", #<<
  spatiotemporal = "iid", #<<
  silent = FALSE # show progress!
)
```

---

# Inspecting the model convergence

.small[
```{r sanity1, eval=FALSE, echo=TRUE}
sanity(fit)
#> ✔ Non-linear minimizer suggests successful convergence
#> ✔ Hessian matrix is positive definite
#> ✔ No extreme or very small eigenvalues detected
#> ✔ No gradients with respect to fixed effects are >= 0.001
#> ✔ No fixed-effect standard errors are NA
#> ✔ No standard errors look unreasonably large
#> ✔ No sigma parameters are < 0.01
#> ✔ No sigma parameters are > 100
#> ✔ Range parameter doesn't look unreasonably large
```
]

---

# Inspecting the model fit

.small[
```{r print-pcod-st-fit, eval=FALSE, echo=TRUE}
fit
```

```{r pcod-st-print2, echo=TRUE, eval=TRUE}
#> Spatiotemporal model fit by ML ['sdmTMB']
#> Formula: density ~ s(depth) + 0 + as.factor(year)
#> Time column: "year"
#> Mesh: mesh
#> Data: pcod
#> Family: tweedie(link = 'log')
#>                     coef.est coef.se
#> as.factor(year)2003     1.79    0.29 #<<
#> as.factor(year)2004     2.36    0.27 #<<
#> ...
#> 
#> Dispersion parameter: 10.78
#> Tweedie p: 1.49
#> Matern range: 12.85 #<<
#> Spatial SD: 1.81
#> Spatiotemporal SD: 1.78 #<<
#> ML criterion at convergence: 6246.433
```
]

---

# Model residuals

.xsmall[
Warning: these residuals might look off even if the model is fine. Also see `dharma_residuals()` or MCMC residuals. See the ['Residual checking' vignette](https://pbs-assess.github.io/sdmTMB/articles/residual-checking.html).
]

```{r resid1, echo=TRUE, out.width='500px', fig.width=5}
set.seed(1)
rq_res <- residuals(fit) # randomized quantile residuals
qqnorm(rq_res);qqline(rq_res)
```

---

<!-- TODO: slides on other types of residuals -->

# Model residuals in space

.small[
```{r resid2, echo=TRUE, out.width='500px', fig.width=5}
pcod$resids <- residuals(fit)
filter(pcod, year %in% c(2015, 2017)) %>% 
  ggplot(aes(X, Y, colour = resids)) + #<<
  geom_point() +
  facet_wrap(~year) +
  scale_colour_gradient2() +
  coord_fixed()
```
]

---

# Predicting on the survey grid

```{r pcod-st-pred, echo=TRUE}
p <- predict(fit, newdata = qcs_grid)
```

* `est`: Overall estimate in link space (log here)
* `est_non_rf`: Estimate of non-random-field components
* `est_rf`: Estimate of random-field components
* `omega_s`: Spatial random field
* `epsilon_st`: **Spatiotemporal random field**

---

# Plotting overall predictions

.xsmall[
```{r pcod-st-plot-est, echo=TRUE, fig.width=7}
ggplot(p, aes(X, Y, fill = exp(est))) +
  geom_raster() +
  scale_fill_viridis_c(trans = "sqrt") +
  facet_wrap(~year) +
  coord_fixed()
```
]

---

# Plotting overall predictions (truncated)

.xsmall[
```{r pcod-st-plot-est2, echo=TRUE}
max_est <- quantile(p$est, 0.999)
p <- p %>% mutate(est_trim = ifelse(est > max_est, max_est, est))

```

```{r pcod-st-plot-est3, echo=FALSE, fig.width=7}
ggplot(p, aes(X, Y, fill = exp(est_trim))) +
  geom_raster() +
  scale_fill_viridis_c(trans = "sqrt") +
  facet_wrap(~year) +
  coord_fixed()
```
]

---

# Plotting main effect contributions

.xsmall[
```{r pcod-st-plot-non-rf, echo=TRUE, fig.width=7}
ggplot(p, aes(X, Y, fill = exp(est_non_rf))) +
  geom_raster() +
  scale_fill_viridis_c(trans = "sqrt") +
  facet_wrap(~year) +
  coord_fixed()
```
]

---

# Plotting spatial random effects

.xsmall[
```{r pcod-st-plot-omega, echo=TRUE, fig.width=7}
ggplot(p, aes(X, Y, fill = omega_s)) +
  geom_raster() +
  scale_fill_gradient2() +
  facet_wrap(~year) +
  coord_fixed()
```
]

---

# Plotting spatiotemporal random effects

.xsmall[
```{r pcod-st-plot-eps, echo=TRUE, fig.width=7}
ggplot(p, aes(X, Y, fill = epsilon_st)) +
  geom_raster() +
  scale_fill_gradient2() +
  facet_wrap(~year) +
  coord_fixed()
```
]

---

# Visualizing spatial uncertainty

By sampling from the joint precision matrix

.small[
```{r pcod-st-sims, echo=TRUE}
psims <- predict(fit, newdata = qcs_grid, nsim = 500) #<<
dim(psims)
psims[1:3, 1:4]

p$sd <- apply(psims, 1, sd) #<<
```
]

---

# Visualizing spatial uncertainty

.small[
```{r pcod-st-sims-plot, echo=TRUE, fig.width=7, out.width="700px", fig.asp=0.5}
filter(p, year %in% c(2013, 2015, 2017)) %>% 
  ggplot(aes(X, Y, fill = sd)) +
  geom_raster() +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10") +
  coord_fixed()
```
]

---




<!-- # Calculating an area-weighted population index -->

<!-- .xsmall[ -->
<!-- ```{r pcod-st-index1, echo=TRUE, fig.width=6, fig.asp=0.5, out.width="600px"} -->
<!-- p2 <- predict(fit, newdata = qcs_grid, return_tmb_object = TRUE) #<< -->
<!-- index <- get_index(p2, area = 4, bias_correct = TRUE) #<< -->
<!-- ggplot(index, aes(year, est)) + geom_line() + -->
<!--   geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) + -->
<!--   xlab('Year') + ylab('Biomass estimate (kg)') -->
<!-- ``` -->
<!-- ] -->

<!-- --- -->

<!-- # Calculating the center of gravity -->

<!-- .small[ -->
<!-- ```{r pcod-st-cog, echo=TRUE, fig.width=6, out.width="600px"} -->
<!-- cog <- get_cog(p2, format = "wide") #<< -->
<!-- ggplot(cog, aes(est_x, est_y, colour = year)) +  -->
<!--   geom_linerange(aes(xmin = lwr_x, xmax = upr_x)) + -->
<!--   geom_linerange(aes(ymin = lwr_y, ymax = upr_y)) + -->
<!--   coord_equal() -->
<!-- ``` -->
<!-- ] -->
