---
title: "Priors and parameter bounds"
subtitle: "DFO TESA sdmTMB workshop"
author: ""
institute: ""
date: "January 16--19 2023"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

<!-- Build with: xaringan::inf_mr() -->

```{r preamble, include=FALSE, cache=FALSE}
source(here::here("dfo-tesa-2023/preamble.R"))
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(dplyr)
library(sdmTMB)
library(ggplot2)
```

# Can maximum likelihood have priors!? 

* Penalized maximum likelihood estimation

* Maximize 

$$L(\theta | y) \cdot P(\theta | \alpha)$$
* $\theta$ are parameters
* $y$ are data
* $\alpha$ are hyperparameters controlling prior distribution  

---

# Penalized estimation in sdmTMB


```{r echo = TRUE, eval=FALSE}
fit <- sdmTMB(
  ...,
  priors = sdmTMBpriors(...), #<<
  ...
)
```

* `?sdmTMBpriors` 

---

# What can we put priors on?

$\beta$ fixed effects  

$\phi$ Observation dispersion parameter (e.g., standard deviation for Gaussian, overdispersion for nbinom2)  

$\rho$ AR(1) parameter for spatiotemporal fields  

$p$ Tweedie parameter (variance power)  

$h$ Matérn range  

$\sigma$ Matérn standard deviation

---

# Example: regression coefficients

* Simple prior on $b_{1}$

* Note: don't have to specify a prior on all coefficients  

.small[
```{r echo=TRUE}
mesh <- make_mesh(pcod,
  xy_cols = c("X", "Y"),
  cutoff = 10
)
fit_prior <- sdmTMB(
  present ~ depth,
  data = pcod, mesh = mesh,
  family = binomial(link = "logit"),
  priors = sdmTMBpriors( #<<
    b = normal(location = c(4.2, NA), scale = c(0.1, NA)) #<<
  ) #<<
)
```
]


---

# Example: regression coefficients

```{r echo=TRUE}
tidy(fit_prior)
```
  
Compare to model without priors  
```{r echo = FALSE}
fit <- sdmTMB(
  present ~ depth,
  data = pcod,
  mesh = mesh,
  family = binomial(link = "logit"),
  spatiotemporal = "off"
)
```

```{r echo=TRUE}
tidy(fit)
```

---

# Example: regression coefficients

* Covariance in fixed effects can be included with `mvnormal()`

.small[
```{r echo=TRUE, eval=FALSE}
fit_prior <- sdmTMB(
  present ~ depth,
  data = pcod,
  mesh = mesh,
  family = binomial(link = "logit"),
  spatiotemporal = "off",
  priors = sdmTMBpriors(
    b = mvnormal(b_loc, scale = b_Sigma)) #<<
)
```
]

---

# Example: Matérn priors

* Spatial data may not be very informative w.r.t. spatial range and variance

* Penalized complexity (PC) priors may help with model convergence
  * [Fuglstad et al. 2016](https://www.tandfonline.com/doi/full/10.1080/01621459.2017.1415907?casa_token=TcqJrOvGCa4AAAAA%3AbPZ1eP-KoRmks3A-kI4QngKNDBNssNQBEAEQs98wemDSH0B_yZv_1q_f5waRhXVv2-ATm3RpmglH)  
  
* Penalty on complexity? 
  * Simple random field = low variance, infinite range
  * Complexity = increasing variance, small spatial range

---

# PC priors parameterized in terms of thresholds

* $\mathrm{Pr}(h > \texttt{range_gt} ) = 1 - \texttt{range_prob}$
* $\mathrm{Pr}(\sigma < \texttt{sigma_lt} ) = 1 - \texttt{sigma_prob}$
```{r echo = TRUE, eval = FALSE}
pc_matern(
  range_gt = 10, range_prob = 0.05, #<<
  sigma_lt = 3, sigma_prob = 0.05 #<<
)
```

---

# Visualizing PC priors

```{r, echo=TRUE, fig.asp=0.7, out.width="550px"}
sdmTMB::plot_pc_matern(
  range_gt = 1, sigma_lt = 1.5, 
  range_prob = 0.05, sigma_prob = 0.05
)
```

---

# Implementing PC prior: spatial random field

.small[
```{r echo=TRUE, eval=FALSE}
fit_prior <- sdmTMB(
  present ~ depth,
  data = pcod,
  mesh = mesh,
  family = binomial(link = "logit"),
  spatiotemporal = "off",
  priors = sdmTMBpriors(matern_s = pc_matern(
    range_gt = 10, #<<
    sigma_lt = 3, #<<
    range_prob = 0.05, # default #<<
    sigma_prob = 0.05  # default #<<
  ))
)
```
]

---

# Using PC priors with spatiotemporal model

* Is range ( $h$ ) shared? `share_range` argument
  * if so, probably not needed to specify 2 sets of PC priors
  
* For model without shared ranges, can specify separate PC priors

```{r echo = TRUE, eval = FALSE}
priors <- sdmTMBpriors(
  matern_s = pc_matern(...),
  matern_st = pc_matern(...)
)
```


---

# Bounds

* Sometimes we may want to put hard bounds on a parameter
* All optional, lower/upper both need not be specified
* See `?sdmTMBcontrol`
* Example: constrain depth effect on occurrence to be negative

.small[
```{r eval=FALSE, echo=TRUE}
fit <- sdmTMB(
  present ~ depth,
  ...,
  control = sdmTMBcontrol(
    lower = list(b_j = c(NA, NA)), #<<
    upper = list(b_j = c(NA, 0))  #<<
  )
)
```
]
