---
title: "Spatial modeling of presence-only data"
subtitle: "DFO TESA sdmTMB workshop"
author: ""
institute: ""
date: "January 16--19 2023"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "theme.css"]
    lib_dir: libs
    nature:
      highlightLines: true
      countIncrementalSlides: false
---

<!-- Build with: xaringan::inf_mr() -->

```{r preamble, include=FALSE, cache=FALSE}
source(here::here("dfo-tesa-2023/preamble.R"))
do.call(knitr::opts_chunk$set, knitr_opts)
```

```{r libs, include=FALSE}
library(dplyr)
library(sdmTMB)
library(ggplot2)
```

# Spatial patterning of trees: bei dataset

* Barro Colorado Island = classic in ecology
  * Hubbell's (2001) unified neutral theory of biodiversity
  
```{r dat, echo=FALSE}
dat <- data.frame(
  x = spatstat.data::bei$x,
  y = spatstat.data::bei$y
)
```

```{r plot-trees, fig.asp=0.5}
ggplot(dat, aes(x, y)) +
  geom_point(col = "darkblue", alpha = 0.1) +
  coord_cartesian(expand = FALSE)
```

---

# Pseudo-absences  

* Georeferenced locations of trees, no 0s  

* Need to generate 0s: how?
  * quadrature points ([Renner et al. 2015](https://doi.org/10.1111/2041-210X.12352))  
  
* Strategy?
  * generate regularly spaced or random?
  * Large enough so that predictive performance does not change as more are added

---
  
# Pseudo-absences from sdmTMB

* Example of uniform grid strategy

```{r make-zeros, echo = TRUE}
res <- 5
zeros <- expand.grid(
  x = seq(0, 1000, by = res),
  y = seq(0, 500, by = res)
)
```

---

# Bind the observed and pseudo-zeros together

.small[
```{r bind-dat, echo=TRUE}
dat$present <- 1
zeros$present <- 0
all_dat <- rbind(dat, zeros)

mesh <- make_mesh(
  all_dat,
  xy_cols = c("x", "y"),
  cutoff = 25 # min. distance
)

mesh$mesh$n # extract number of vertices/knots
```
]

---

# Combined data 

* blue dots are data; red grid dots are quadrature points
* grey triangles are from the SPDE mesh

.small[
```{r mesh-viz, fig.asp=0.5, echo=FALSE}
all_dat$fpres <- as.factor(all_dat$present)
ggplot() +
  inlabru::gg(mesh$mesh) +
  geom_point(
    data = all_dat,
    aes(x = x, y = y, col = fpres), size = 0.2, alpha = 0.3
  ) +
  coord_equal() +
  # guides(colour = guide_legend(override.aes = list(alpha = 1))) + 
  guides(col = guide_legend(title = "Present"))
```
]

---

# Infinitely Weighted Logistic Regression (IWLR)

* [Fithian & Hastie (2013)](https://doi.org/10.1214/13-AOAS667)

```{r make-iwlr, echo=TRUE}
nW <- 1.0e6
all_dat$wt <- nW^(1 - all_dat$present)
```

* weights can be passed into model of choice
  * `glm()`, `glmmTMB()`, etc.
  * adding random fields makes this a "spatial log-Gaussian Cox process"

---

# IWLR & sdmTMB

* convergence may be affected by size of pseudo-absences    
* intercept and log-likelihood affected by `nW`
```{r iwlr-fit, echo = TRUE}
fit <- sdmTMB(
  present ~ 1,
  data = all_dat,
  mesh = mesh,
  family = binomial(link = "logit"),
  weights = all_dat$wt #<<
)
```

---

# Downweighted Poisson Regression (DWPR)

* Similar to IWLR, uses different weights
* Doesn't have same arbitrary effects on intercept, likelihood  

```{r make-dwpr, echo=TRUE}
# small values at presence locations
all_dat$wt <- 1e-6

# pseudo-absences: area per quadrature point
tot_area <- diff(range(dat$x)) * diff(range(dat$y))
n_zeros <- length(which(all_dat$present == 0))

all_dat$wt <- ifelse(all_dat$present == 1,
  1e-6, tot_area / n_zeros
)
```

---

# DWPR & sdmTMB

```{r fit-dwpr, echo=TRUE}
fit <- sdmTMB(
  present / wt ~ 1,
  data = all_dat,
  mesh = mesh,
  family = poisson(link = "log"),
  weights = all_dat$wt #<<
)
```

---

# Plotting spatial random effects

.small[
```{r plot-rf, echo=FALSE}
p <- predict(fit, newdata = zeros)
ggplot(p, aes(x, y, fill = omega_s)) +
  geom_raster() +
  scale_fill_gradient2() +
  coord_fixed(expand = FALSE)
```
]

---

# Predictions in link (log) space

.xsmall[
```{r plot-link, echo=FALSE}
ggplot(p, aes(x, y, fill = est)) +
  geom_raster() +
  scale_fill_viridis_c() +
  coord_fixed(expand = FALSE) # +
# geom_point(aes(x, y, size = n),
# data = dat, pch = 21, colour = "grey60", inherit.aes = FALSE)
```
]

---

# Does adding more pseudo-absences improve performance?

* Increase 0s from ~ 20K to 30K

* AUC similar (other criteria could be used) 

```{r inc-points-auc, echo=FALSE}
dat <- data.frame(
  x = spatstat.data::bei$x,
  y = spatstat.data::bei$y
)

res <- 4
zeros <- expand.grid(
  x = seq(0, 1000, by = res),
  y = seq(0, 500, by = res)
)

dat$present <- 1
zeros$present <- 0
all_dat <- rbind(dat, zeros)
mesh <- make_mesh(
  all_dat,
  xy_cols = c("x", "y"),
  cutoff = 25 # min. distance between knots in X-Y units
)
# mesh$mesh$n # extract number of vertices/knots

all_dat$wt <- 1e-6

# pseudo-absences: area per quadrature point
area <- diff(range(dat$x)) * diff(range(dat$y))
all_dat$wt[which(all_dat$present == 0)] <-
  area / length(which(all_dat$present == 0))

fit2 <- sdmTMB(
  present / wt ~ 1,
  data = all_dat,
  mesh = mesh,
  family = poisson(link = "log"),
  weights = all_dat$wt
)

all_dat$p <- predict(fit2)$est # total predictions, logit
rocr <- ROCR::prediction(plogis(all_dat$p), all_dat$present)
ROCR::performance(rocr, measure = "auc")@y.values[[1]]
```

---

# What about using a higher resolution mesh?

* Change cutoff from 25 to 15

* Knots change from ~700 to ~1750

* Marginal gains in AUC with finer mesh
  * Note: it's not adding more pseudo-absences but changing the mesh that's more important here  
  
```{r auc-highres-mesh, echo=FALSE}
dat <- data.frame(
  x = spatstat.data::bei$x,
  y = spatstat.data::bei$y
)

res <- 4
zeros <- expand.grid(
  x = seq(0, 1000, by = res),
  y = seq(0, 500, by = res)
)

dat$present <- 1
zeros$present <- 0
all_dat <- rbind(dat, zeros)
mesh <- make_mesh(
  all_dat,
  xy_cols = c("x", "y"),
  cutoff = 15 # min. distance between knots in X-Y units
)
# mesh$mesh$n # extract number of vertices/knots

all_dat$wt <- 1e-6

# pseudo-absences: area per quadrature point
area <- diff(range(dat$x)) * diff(range(dat$y))
all_dat$wt[which(all_dat$present == 0)] <-
  area / length(which(all_dat$present == 0))

fit3 <- sdmTMB(
  present / wt ~ 1,
  data = all_dat,
  mesh = mesh,
  family = poisson(link = "log"),
  weights = all_dat$wt
)

all_dat$p <- predict(fit3)$est # total predictions, logit
rocr <- ROCR::prediction(plogis(all_dat$p), all_dat$present)
ROCR::performance(rocr, measure = "auc")@y.values[[1]]
```

---

# Benefits of pseudo-absence modeling

* Estimate of spatial range isn't sensitive to choice of raster / lattice resolution
* Doesn't require wrangling raw data (e.g. aggregating to larger cell size to model counts)
